\chapter{Single parameter regression}

\section{Mean only regression}



Consider least squares where we only want horizontal lines. 
Let our outcome be
$
\by = (y_1,\ldots, y_n)^t
$ and recall that $\bone_n$ is an $n$ vector of ones. We want
to minimize 
$f(\mu) = ||\by - \mu \bone||^2$ with respect to $\mu$. 

Taking derivatives with respect to $\mu$ we obtain that
$$
\frac{d f}{d\mu} = - 2n \bar y  + 2 n \mu.
$$
This has a root at $\hat \mu = \bar y$. Note that the second
derivative is $2n>0$. 
Thus, the average is the least squares estimate in the
sense of minimizing the Euclidean distance between the
observed data and a constant vector. We can think of this
as projecting our $n$ dimensional onto the best $1$
dimensional subspace spanned by the vector $\bone$. We'll
rely on this form of thinking a lot throughout the text.

\section{Coding example}
\href{}{}

Let's use the \texttt{diamond} dataset 
\begin{verbatim}
> library(UsingR); data(diamond)
> y = diamond$price; x = diamond$carat
> mean(y)
[1] 500.0833
> #using least squares
> coef(lm(y ~ 1))
[1] 500.0833
\end{verbatim}

Thus, in this example the mean only least squares estimate obtained via \texttt{lm} 
is the empirical mean.

\section{Regression through the origin}

\href{https://www.youtube.com/watch?v=1ZFED8AcHWc&index=7&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y}{Watch this video before beginning.}

Let $\bx = (x_1,\ldots, x_n)'$ be another vector. Consider now the
regression through the origin problem. We want to minimize
$f(\beta) = ||\by - \beta \bx||^2$ with respect to $\beta$.
This is called regression through the origin for the following
reason. First note that the pairs, $(x_i, y_i)$,
form a scatterplot. Least squares is then finding the best
multiple of the $\bx$ vector to approximate $\by$. That is,
finding the best line of the form $y = \beta x$ to fit
the scatter plot. Thus we are considering lines through the origin
hence the name regression through the origin.

Notice that
$
f(\beta) = \by^t \by - 2 \by^t \bx + \bx^t \bx.
$
Then 
$$
\frac{df}{d\beta} = -2\by'\bx  + 2 \bx^t \bx 
$$
Setting this equal to zero we obtain the famous equation:
$$
\hat \beta = \frac{\by^t \bx}{\bx^t \bx}
=\frac{\ip{\by}{\bx}}{\ip{\bx}{\bx}}
$$
We'll leave it up to the reader to check the second derivative 
condition. Also, we'll leave it up to you to show that the mean 
only regression is a  special case that agrees with the result.

Notice that we have shown the function
$$
g : \mathbb{R}^n \rightarrow \mathbb{R}
$$
defined by $g(\by) = \frac{\ip{\by}{\bx}}{\ip{\bx}{\bx}}\bx$ projects
any $n$ dimensional vector $\by$ into the linear space spanned
by the single vector $\bx$, $\{\beta \bx ~|~ \beta \in \mathbb{R}\}$.

\section{Centering first}

\href{https://www.youtube.com/watch?v=1ss_FYtiSHo&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=8}{Watch this video before beginning.}

A line through the origin is often not useful. Consider centering the
$\by$ and $\bx$ first. The the origin would be at the mean of the
$\by$ vector and the mean of the $\bx$ vector. Let
$\tilde \by = \left\{ \bI - \bone_n (\bone_n^t \bone_n)^{-1} \bone_n^t \right\} \by$ and $\tilde \bx = \left\{ \bI - \bone_n (\bone_n^t \bone_n)^{-1} \bone_n^t \right\} \bx$. Then regression through the origin (minimizing
$||\tilde \by - \gamma \tilde \bx||^2$ for $\gamma$)
for the centered data yields
the solution 
$
\hat \gamma = \frac{\ip{\tilde \by}{\tilde \bx}}{\ip{\tilde \bx}{\tilde \bx}}.
$
However, from the previous chapter, we know that
$$
\ip{\tilde \by}{\tilde \bx} = \by^t \left\{ \bI - \bone_n (\bone_n^t \bone_n)^{-1} \bone_n^t \right\} \bx^t
= (n-1)\hat{\rho}_{xy} \sigma_x \sigma_y
$$
and similarly $\ip{\tilde \bx}{\tilde \bx} = (n-1)\hat{\sigma}_y^2$
where $\hat \rho_{xy}$ and $\hat{\sigma}_y^2$ are the empirical correlation and
variance, respectively. Thus our regression through the origin estimate is
$$
\hat \gamma = \rho_{xy} \frac{\sigma_y}{\sigma_x}.
$$
Thus the best fitting line that has to go through the center of the data
has a slope equal to the correlation times the ratio of the standard deviations.
If we reverse the role of $\bx$ and $\by$, we simply invert the ratio of the
standard deviations. Thus we also note, that if we center and scale our
data first so that the resulting vectors have mean 0 and variance 1, our
slope is exactly the correlation between the vectors.

\subsection{Coding example}

\href{https://www.youtube.com/watch?v=CrqNQEYF-nU&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=9}{Watch this video before beginning.}

Let's continue with the diamond example. We'll center the variables first.
\begin{verbatim}
> yc = y - mean(y);
> xc = x - mean(x)
> sum(yc * xc) / sum(xc * xc)
[1] 3721.025
> coef(lm(yc ~ xc - 1))
      xc 
3721.025 
> cor(x, y) * sd(y) / sd(x)
[1] 3721.025
\end{verbatim}

\section{Bonus videos}

\noindent
Watch these videos before moving on. (I had created them beofre I reorganized chapters.)

\href{https://www.youtube.com/watch?v=lmv88DtCNiU&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=10}{Sneak preview of projection logic.}

\href{https://www.youtube.com/watch?v=0Ld7sZ8FUs0&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=11}{Coding example.}

\href{https://www.youtube.com/watch?v=U5FAOdBDb90&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=12}{Sneak preview of linear regression.}

\href{https://www.youtube.com/watch?v=Ir1L-STFKfA&list=PLpl-gQkQivXhdgUCdaUQcdb31CRe8Mm2y&index=13}{Sneak preview of regression generalizations.}