\section{Bases}


Recall that any set of linearly independent vectors forms a basis,
specifically for the space spanned by linear combinations. Therefore,
least squares is projecting our data into the space created by
the basis defined by the columns of our design matrix. Of note,
certain bases are of particular importance as the spaces that
they create are contextually meaningful for many scientific
problems. The most notable example are Fourier bases. In this
case, we project our data into the space spanned by harmonic
functions. In problem where the study of periodicities is 
of interest, this is of tremendous use.

\subsection{Introduction to full rank bases}


When our $\bX$ matrix is $n\times n$ of rank $n$, then the least squares
fitted values $\bX (\bX' \bX)^{-1} \bX' \by = \bX \bbeta = \hat \by$ is simply a linear reorganization
of $\by$ as it's projecting it from $\mathbb{R}^n$ to $\mathbb{R}^n$. 
Despite not summarizing $\by$ in any meaningful way, this is often
a very meaningful thing to do, particularly when the basis is orthonormal.
This full rank linear transformation of $\by$ is simply called a 
``transform''. Notable bases then get named as their name then ``transform''. 
The best examples include the Fourier transform and the Wavelet transform.
Often, because we're applying the transform to vectors with discrete
indices, rather than continuous functions, the label ``discrete'' is
affixed, such as the discrete Fourier transform. Looking back to
our Hilbert space discussion from earlier the extension to continuous
spaces is conceptually straightforward. 

Let $\bX = [\bx_1 ~ \ldots ~ \bx_n]$ be our basis, so that 
$\xtx = I$. In this case note that 
$$
\hat \bbeta
= \bX' \by 
= 
\left(
\begin{array}{c}
 \ip{\bx_1}{\by} \\
 \vdots \\
 \ip{\bx_n}{\by}
\end{array}
\right).
$$
Thus, our coefficients are exactly the inner products of the basis elements (columns of $\bX$) and the outcome. Our fitted values are
$$
\hat \by = \bX \hat \bbeta = \sum_{i=1}^n \bx_i \ip{\bx_i}{\by}.
$$
Consider deleting columns from $\bX$, say 
$$
\bW = [\bx_{i_1} \ldots \bx_{i_p}]
$$
and minimizing $||\by - \bW \bgamma||^2$. Since $\bW$ is also orthonormal (but not full rank) we get that 
$$
\hat \bgamma = \sum_{j=1}^p \ip{\bx_{i_j}}{\by} ~~~
\mbox{and} ~~~
\hat \by = \sum_{j=1}^p \bx_{i_j} \ip{\bx_{i_j}}{\by}
=  \sum_{j=1}^p \bx_{i_j} \beta_{i_j}.
$$
That is, the coefficients from the model with columns removed are identical
to the coefficients from the full model. Transforms are often then
accomplished by getting the full transform and using a subset of the 
coefficients to get the fitted values. 


\subsection{Principal component bases}


One orthonormal basis is always at our disposal, the principal component basis. 
Consider the case where $\bX$ has $p > n$ columns of full row rank.
Let $\bX = \bU \bD \bV'$ be the singular value decomposition of $\bX$ so
that $\bU$ is $n \times n$, so that, $\bU' \bU = \bI$ and $\bV'$
is $n\times p$ so that $\bV' \bV = \bI$ and $\bD$ is diagonal containing
$n$ singular values. The matrix $\bU$ is a full rank version of the
column space of $\bX$. Notice that minimizing
$$
||\by - \bX \bbeta||^2
$$
has $n$ equations and $p>n$ unknowns and that
$$
||\by - \bX \bbeta||^2
= 
||\by -  \bU \bD \bV' \bbeta||^2
$$
So that by defining $\bgamma = \bD \bV' \bbeta$ and minimizing 
$$
||\by -  \bU \bgamma||^2.
$$
we have a full rank design matrix created out of the columns of $\bX$ since
$\bU = \bX \bV \bD^{-1}$. 
The fitted values are merely $\bU' \by$.

Note, if $\bX$ has been centered, then 
$$
\frac{1}{n-1} \bX' \bX = \frac{1}{n-1} \bV \bD^2 \bV'
$$
is the covariance matrix between the columns of $\bX$. Furthermore,
notice that the total variability represented by the trace of the covariance
matrix is, 
$$\frac{1}{n-1} \mbox{tr}(\bX' \bX)
= \frac{1}{n-1} \mbox{tr}(\bV \bD^2 \bV')
=\frac{1}{n-1} \mbox{tr}(\bD^2 \bV' \bV)
= \frac{1}{n-1} \mbox{tr}(\bD^2)
$$
The sum of the squared singular values equals the total variability in the
design matrix. 
The singular
vectors and values are typically ordered in the terms of decreasing
variability. Therefore, keeping only a few of them represents a dimension
reduction that preserves the greatest amount of variability.

Thus, we once one calculates $\bU' \bY$ we have all possible submodel
fits of the columns of $\bU$, where $\bU$ is an meaningful summary
of $\bX$. Typically one takes a few of the first columns of $\bU$
so that the related eigenvalues explain a large proportion of the
total variability. We haven't discussed an intercept. However, 
one usually mean centers $\bY$ and $\bX$ first.


\subsection{Coding example}

The following code goes through calculation of SVD and eigenvalue 
decompositions.

\begin{verbatim}
data(swiss)
y = swiss$Fertility
x = as.matrix(swiss[,-1])
n = nrow(x)
decomp = princomp(x, cor = TRUE)
plot(cumsum(decomp$sdev^2) / sum(decomp$sdev^2), type = "l")
decomp2 = eigen(cor(x))
xnorm = apply(x, 2, function(z) (z - mean(z)) / sd(z))
decomp3 = svd(xnorm)
round(rbind(decomp2$vectors, decomp$loadings, decomp3$v),3)
round(rbind(decomp2$values, decomp$sdev^2, decomp3$d ^ 2 / (n - 1)), 3)
plot(decomp3$u[,1], decomp$scores[,1])
plot(decomp3$u[,1], xnorm %*% decomp2$vectors %*% diag(1 / sqrt(decomp2$values))[,1])
\end{verbatim}

