\chapter{Distributional results}
In this chapter we assume that $\bY \sim N(\bX \bbeta, \sigma^2 \bI)$. 


\section{Quadratic forms}
\label{sec:qf}
Let $\bA$ be a symmetric matrix (not necessarily full rank)
and let $\bX \sim N(\bmu, \bSigma)$. Then
$(\bX - \bmu)^t \bA (\bX - \bmu) \sim \chi^2_{p}$ if 
$\bA \bSigma$ is idempotent and $p=\mbox{Rank}(\bA)$. 

As an example, note that $(\bX - \bmu)^t \bSigma^{-1} (\bX - \bmu)$
is clearly Chi-squared($n$). This is most easily seen by the fact that
$(\bX - \bmu)\bSigma^{-1/2} = Z$ is a vector of iid standard normals
and thus the quadratic form is the sum of their squares. Using
our result, $\bA = \bSigma^{-1}$ in this case and  
$\bA \bSigma = \bI$, which is idempotent. The rank of $\bSigma^{-1}$ is $n$.


Let's prove our result. Let $\bA = \bV \bD^2 \bV^t$ where $\bD$ is diagonal (with $p$ non zero entries)
and $\bV \bV^t = \bI$. The assumption of idempotency gives us that
$\bA \bSigma \bA \bSigma = \bA \bSigma$. Pluggging in our decomposition for $\bA$
and using the orthonormality of the columns of $\bV$ we get that
$\bD \bV^t \bSigma \bV \bD = \bI$. Then note that
\begin{equation}
\label{eq:quad}
(\bX - \bmu)^t \bA (\bX - \bmu) = (\bX - \bmu)^t \bV \bD  \bD \bV^t (\bX - \bmu).
\end{equation}
But $\bD \bV^t (\bX - \bmu) \sim N(\bzero, \bD \bV^t \bSigma \bV \bD)$, 
which has variance equal to $\bI$. Thus in Equation \eqref{eq:quad} we have
the sum of $p$ squared iid standard normals and is thus Chi-squared $p$. 

\section{Statistical properties}
For homework, show that $\hat \bbeta$ is normally distributed with moments: 

$$
E[\hat \bbeta] = \bbeta ~~~\mbox{and}~~~ \Var(\hat \bbeta) = \xtxinv \sigma^2.
$$

The residual variance estimate is $S^2 = \frac{1}{n-p}\be^t \be$. Using Section
\ref{sec:qfm} we see that it is unbiased, $E[S^2] = \sigma^2$.  
Note also that:
\begin{eqnarray*}
\frac{n-p}{\sigma^2} S^2 & = & \frac{1}{\sigma^2}\by^t (\bI - \hatmat) \by \\
& = & \frac{1}{\sigma^2}(\by - \bX \bbeta)^t \{\bI - \hatmat\} (\by - \bX \bbeta)
\end{eqnarray*}
is a quadratic form of as discussed in Section \eqref{sec:qf}.  
Furthermore
$$
\frac{1}{\sigma^2}\{\bI - \hatmat\} \Var(Y) = \{\bI - \hatmat\},
$$
which is idempotent. For symmetric idempotent matrices, the rank equals
the trace; the latter of which is easily calculated as
$$
\mbox{tr}\{\bI - \hatmat\} = \mbox{tr}\{\bI\} - \mbox{tr}\{\hatmat\}
= n - \mbox{tr}\{\xtxinv \xtx\} = n - p.
$$
Thus, $\frac{n-p}{\sigma^2} S^2$ is Chi-squared $n-p$. The special
case of this where $\bX$ has only an intercept yields the usual
empirical variance estimate.


\section{T statistics}
We can now develop T statistics.
Consider the linear contrast $\hat \bbeta^t \bt$. First note that
$\hat \bbeta^t \bt$ is $N(\bbeta^t \bt, \bt^t \xtxinv \bt \sigma^2)$. 
Furthermore,
$\Cov(\hat \bbeta, \be) = \Cov(\xtxinv \bX^t \bY, \{\bI - \hatmat\} \bY) = 0$
since $\xtxinv \bX^t \{\bI - \hatmat\} = \bzero$. Thus the residuals
and estimated coefficients are independent, implying that $\hat \bbeta^t \bt$
and $S^2$ are independent.
Therefore,
$$
\frac{\hat \bbeta \bt - \bbeta \bt}{\sqrt{\bt^t \xtxinv \bt \sigma^2}}
/ \sqrt{\frac{n-p}{\sigma^2} S^2 / (n-p)}
=\frac{\hat \bbeta \bt - \bbeta \bt}{\sqrt{\bt^t \xtxinv \bt S^2}}
$$
is a standard normal divided by the square root of an independent Chi-squared
over its degrees of freedom, thus is $T$ distributed with $n-p$ degrees of freedom.

\section{F tests}
Consider testing the hypothesis that $H_0:\bK \bbeta = 0$ versus not equal for $\bK$
of full row rank (say $v$). 
Notice that $\bK \hat \bbeta \sim N(\bK \bbeta, \bK \xtxinv \bK^t \sigma^20$ and thus
$$
(\bK \hat \bbeta - \bK \bbeta)^t  \{\bK \xtxinv \bK^t \sigma^2 \}^{-1} (\bK \hat \bbeta - \bK \bbeta)
$$
is Chi-squared with $v$ degrees of freedom. Furthermore, it is independent of
$\be$ being a function of $\hat \bbeta$. Thus:
$$
(\bK \hat \bbeta - \bK \bbeta)^t  \{\bK \xtxinv \bK^t\}^{-1} (\bK \hat \bbeta - \bK \bbeta) / v S^2
$$
forms the ratio of two independent Chi-squared random variables over their degrees of freedom, which is an F distribution.







